% This LaTeX document was generated using the LaTeX backend of PlDoc,
% The SWI-Prolog documentation system



\section{xml.pl -- xml parsing}

\label{sec:xml}

\file{xml.pl} : Contains xml_parse/[2,3] a bi-directional XML parser written in
Prolog.

Copyright (C) 2001, 2002 Binding Time Limited

Current Release: 1.5

TERMS AND CONDITIONS:

This program is offered free of charge, as unsupported source code. You may
use it, copy it, distribute it, modify it or sell it without restriction.

We hope that it will be useful to you, but it is provided "as is" without
any warranty express or implied, including but not limited to the warranty
of non-infringement and the implied warranties of merchantability and fitness
for a particular purpose.

Binding Time Limited will not be liable for any damages suffered by you as
a result of using the Program. In no event will Binding Time Limited be
liable for any special, indirect or consequential damages or lost profits
even if Binding Time Limited has been advised of the possibility of their
occurrence. Binding Time Limited will not be liable for any third party
claims against you.

History:
\$Log: \file{xml.pl},v \$
Revision 1.2 2002-05-25 23:17:54+01 john
Set Current Release to 1.5.

Revision 1.1 2002-01-31 21:04:45+00 john
Updated Copyright statements.

Revision 1.0 2001-10-17 20:46:24+01 john
Initial revision\vspace{0.7cm}

\begin{description}
    \predicate[det]{xml2pl}{2}{+Input, -Output}
Create and Prolog representation, \arg{Output}, of the XML file with name \arg{Input}
added by Neng-Fa Zhou 

    \predicate{pl2xml}{2}{+Document, -Codes}
Convert Prolog representation to XML document
\end{description}

